---
title: "ElectionOutcomes"
author: "Sophia Awoyinka, Angie Liu, Golbon Marandiz"
format: 
  html:
    embed-resources: true
editor: visual
---

```         
```

# Question

How does monetary contribution affect election outcomes?

# Introduction

This project investigates the relationship between the economic performance of states (measured as GDP per capita) and monetary contributions to political campaigns, examining their impact on election outcomes. By analyzing state-level GDP data and detailed candidate financial information, the goal is to uncover potential correlations between economic affluence and the success of political candidates.

# Data Sources

## Dataset 1

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)

funding_data <- read_csv("CandidateInfo.csv")

gdp_data <- read_csv("gdp_per_capita_by_state.csv")

funding_data$total_funds <- gsub("[$,]", "", funding_data$total_funds)

funding_data <- funding_data |>
  select(id, name, state, affiliation, total_funds, winner, votes)

gdp_data <- gdp_data |>
  select(state, year2016)

glimpse(funding_data)

```

\*Source: [**GDP Per Capita by State**](https://www.kaggle.com/datasets/solorzano/gdp-per-capita-in-us-states) **-** Description:\*\* This dataset contains the GDP per capita for each U.S. state from 2013 to 2017. It provides a measure of the economic output of states, which serves as an indicator of their wealth and economic standing.

```{r}
funding_data <- funding_data |>
  drop_na()

ggplot(data = funding_data, aes(x = total_funds, y = votes)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Funds vs Votes recieved", x = "funds", y = "Votes Recieved", alt = "Scatterplot showing relationship between funds and votes recieved. No visible correlation.")
```

**Caption:** This scatter plot visualizes the relationships between the funds raised by candidates in the 2016 congressional election results(x-axis) and the votes received by candidates in an election. Each point represents a candidate.

**Description:** The funds raised span a wide range not explicitly represented on the x-axis. The scatter of the points do not form a clear liner pattern which indicates a weak or indirect correlation between campaign funds and the numbre of votes received. The outliers are the candidates who received significantly higher votes. The weak correlation suggests that funding may not guarantee outcome of congressional elections.

```{r}
funding_data <- funding_data |>
  mutate(funding_range = case_when(
    total_funds > 0 & total_funds <= 30000000 ~ "0-30M",
    total_funds > 30000000 & total_funds <= 60000000 ~ "30M-60M",
    total_funds > 60000000 ~ "60M+"))

glimpse(funding_data)
unique(funding_data$winner)
```

```{r}
ggplot(funding_data, aes(x = funding_range, fill = winner)) +
  geom_bar(position = "stack") +  # Stacked bar with proportions
  labs(
    title = "Proportion of Election Results by Funding Range",
    x = "Funding Range",
    y = "Proportion of Candidates",
    fill = "Win",
    alt = "Bar graph showing correlation between successful elections and how much finding winners recieve."
  ) +
  scale_fill_manual(
    values = c("Y" = "green"),  
    na.value = "gray" 
  ) +
  theme_minimal()
```

**Caption:** This graph illustrates whether states with higher economic performance contribute more significantly to political campaigns.The cateogories on th x-axis range are the funding ranges were are 0-30M,30-60M, 60M+ and NA.

**Analysis:** The bar chart titled "Proportion of Election Results by Funding Range" indicates a nuanced relationship between campaign funding levels and electoral success.The candidates in the lowest funding range of the data set exhibit a higher proportion of desired election outcomes compared to those in the30-60M range suggesting diminishing returns in mid-range campaigns. The highest range of 60M+ are heavily funded however the data does not indicate that these candidates experienced a higher win proportion. This graph implies that there is not a direct relationship between campaign funding and election results.

```{r}

#get gdp ranges
gdp_data <- gdp_data |>
  mutate(gdp_range = case_when(
    year2016 > 0 & year2016 <= 40000 ~ "0-40k",
    year2016 > 30000 & year2016 <= 60000 ~ "40k-70k",
    year2016 > 60000 ~ "70k+"))

merged_data <- left_join(gdp_data, funding_data, by = "state")

merged_data <- merged_data |>
  select(state, total_funds, gdp_range, year2016)

head(merged_data, 30)
```

```{r}
get_mode <- function(x) {
  unique_x <- unique(x)
  unique_x[which.max(tabulate(match(x, unique_x)))]
}

by_state <- merged_data |> 
  group_by(state) |> 
  summarise(
    total_funds_sum = sum(as.numeric(total_funds), na.rm = TRUE),
    range = get_mode(gdp_range),
    average_funds_per_range = mean(as.numeric(total_funds), na.rm = TRUE),
    state_gdp = get_mode(year2016),
    .groups = "drop"
  )
head(by_state, 30)
```

```{r}
ggplot(by_state, aes(x = range, y = average_funds_per_range)) +
  geom_bar(stat = "identity", fill = "blue") +  # Create bars with specified color
  labs(
    title = "Avg Funds by GDP Range",
    x = "GDP Range",
    y = "Total Funds Sum"
  ) +
  theme_minimal() +  # Apply a minimal theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Caption:** This graph illustrates the relationship between GDP ranges and the total funds assoicated with them. The ranges are 0-40K, 40-70K, and 70k+ on the x axis and th y axis it the sum of total funds in monetary terms for each GDP range.

**Analysis:** The mid-range of state GPD has the highest average total funds, significantly surpassing the other two ranges. The 0-40k GDP range has a slightly higher sum of total funds than the 70k+ GDP range states. This may suggest that there may not a direct coorelation between GDP and campaign funds as higher GDP could imply more wealth and potentially higher funding, which is not indicative of the dataset. The significantly high total funds in the middle range may warrant further exploration to understand if there are specific outliers skewing these results.

```{r}
ggplot(by_state) +
  geom_histogram(aes(x = total_funds_sum), fill = "orange", color = "black", bins = 30) + 
  labs(
    title = "Distribution of GDP and Total Funding",
    x = "State GDP",
    y = "Total funding given to candidates"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 10)
  ) 
```

**Caption:** This visualization highlights the potential influence of campaign funding and economic affluence on election outcomes.

**Analysis** The "Avg Funds by GDP Range" visualization depicts the distribution of campaign funding in states with GDPs in the 40k–70k range. This could indicate that states within this GDP range are concentrated due to their population size, competitiveness, or economic activity. In contrast, states with GDPs below 40k and above 70k receive comparatively less funding, highlighting disparities influenced by economic and demographic factors.

# Conclusion

This analysis investigates the critical role of campaign funding in electoral success while also revealing the disparities across funding levels and economic performance. The data suggests a nuanced relationship between the economy and electoral results. Although higher funding levels correlate with increases chances of success, there are no linear relationships. Also, states within the mid range of GDP ranges disproportionately attract campaign funds, which can indicate inequalities in political financing. Findings of similar datasets and studies are inconsistent. Some have found a correlation between funding and electoral results. For example, a quantitative study that investigated campaign funding and election results based on 2018-2021 congressional elections show a positive relationship between campaign funding, campaign spending, and electoral outcomes.(Hollibaugh, G. E., & Rothenberg, L. S. (2021). Primo and Milyo(2020) indicate that public perceptions of candidates play more substantial role in shaping public trust, which isn't always gained through highly advertized campaigns.

Future efforts should focus on increasing transparency in government. These steps can create a more equitable and representative political system where financial disparities do not have any effect on electoral outcomes.

# References

1.  **Dataset 1 Source:** [GDP Per Capita by State dataset](https://www.kaggle.com/datasets/solorzano/gdp-per-capita-in-us-state)
2.  **Dataset 2 Source:** [Candidate Info dataset](https://www.kaggle.com/datasets/danerbland/electionfinance?resource=download)
3.  Additional Sources:
    -   Primo, D. M., & Milyo, J. (2020). Campaign finance and American democracy: What the public really thinks and why it matters. University of Chicago Press.

    -   Hollibaugh, G. E., & Rothenberg, L. S. (2021). Money in politics: How does it affect public trust in government? SAGE Open, 11(2), 1–12. https://doi.org/10.1177/21582440211027965
